Variance reduction for MCMC methods via martingale representations

### 2020-03-05 aap/Revision/vr-mcmc-revision-new.tex
lemma 10 update

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
\begin{eqnarray*}
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\[
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\]
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

@@

\begin{lem} 
\label{lem:bias}
Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. Then the squared bias satisfies
\begin{equation}
\label{eq:sq_bias}
\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \mathrm{osc}(f)^2 \left[\gamma + \frac{1}{n}\left(V(x) + \pi_{\gamma}(V)\right)\right]
\end{equation}
where $\lesssim$ means less up to a constant not depending on $n$ and $\gamma$.
\end{lem}

\begin{proof} We can estimate the conditional bias similarly to~\cite[Section~3]{durmus:moulines:2017}:
\begin{equation}
\label{eq:06042018a2}
\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\le
\frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
\tvnorm{\delta_x \kerULA_\gamma^{p}-\pi(\cdot)}^2,
\end{equation}
Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad \tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_2\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
\end{equation}
for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Combining \ref{eq:bias} with the triangle inequality, we obtain \ref{eq:sq_bias}.
\end{proof}

@@

\appendix
\section{Bounds for moments of  ULA}\label{sec:appendix}

We shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the so-called drift condition:
\begin{definition}
\label{def:drift}
We say that the Markov kernel $Q_{\gamma}$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
\begin{eqnarray*}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{eqnarray*}
\end{definition}

\begin{lem}
\label{lem:drift}
 Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel $Q_\gamma$ from \ref{eq:kernel} satisfies drift condition \ref{def:drift} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L^2}$ with drift function $V(x) = 1 + \|x\|^2$, constants $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_1^2}{8} + 2d + m$ with $K_1$ from \ref{eq:H3}, $m$ from {\bf (H2)}.
\end{lem}

\begin{proof} Consider $V(x) = 1 + \|x\|^2$, then
$$
Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}(1+\|y\|^2)\frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{4\gamma}\right)}\,dy =
$$

$$
=1 + \frac{1}{(4\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{4\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_1)$. Note that
$$
\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \gamma^2 \|x - \gamma \nabla U(x)\|^2
$$
and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,2\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = 2\gamma d$. It remains to notice that due to {\bf H1} and {\bf H2} (which implies \ref{eq:H3}),
$$
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq
$$

$$
 \leq (1-\gamma m)\|x\|^2 + 2\gamma^2 L^2 \|x\|^2
$$
Thus, plugging everything into expression for $R_{\gamma}V$ and using $\gamma < \frac{m}{4L}$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + 2\gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (2d + m)
$$
Now let $x \in B(0,K_1)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
R_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + 2d + 2(1+L\gamma)^2\|x\|^2\right) \leq
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_1^2}{8} + 2d + m)
$$
\end{proof}

>>>>

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L < \infty$ such that
\begin{eqnarray*}
\| \nabla U(x)-\nabla U(y) \| \leq L \|x-y\|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K>0$ and $m>0$ such that for any $x\not\in B_{K}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq m\|x\|^2.
\end{equation*}
\end{description}
Let $\pi$ be the probability measure on $\rset^d$ with density $\pi(x)$ of the form \eqref{eq:stationary_distr};
for $\gamma>0$, define the Markov kernel $Q_\gamma$ associated to one step of the ULA algorithm by
\begin{equation}
\label{eq:ula_kernel}
\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y
\end{equation}
for any $A \in \mathcal{B}(\rset^d)$. Note that for a homogeneous Markov chain $\left(X^x_{p}\right)_{p \in \nset_0}$ of the form \eqref{eq:chain} with constant step size $\gamma$, $\mathsf{P}(X_n^x \in A) = \kerULA^n_{\gamma}(x,A)$.

@@

\begin{lem} 
\label{lem:bias}
Let $f$ be a bounded Borel function and $(X^x_{p})_{p \in \nset_0}$ be a Markov chain generated by ULA with constant step size $\gamma$. Then the squared bias satisfies
\begin{equation}
\label{eq:sq_bias}
\left( \mathsf{E}[\rho(f)]-\pi(f)\right)^2 \lesssim \mathrm{osc}(f)^2 \left[\gamma + \frac{1}{n}\left(V(x) + \pi_{\gamma}(V)\right)\right]
\end{equation}
for $V(x) = 1 + \|x\|^2$, where $\lesssim$ means less up to a constant not depending on $n$ and $\gamma$.
\end{lem}

\begin{proof} We can estimate the conditional bias similarly to~\cite[Section~3]{durmus:moulines:2017}:
\begin{equation}
\label{eq:06042018a2}
\left( \mathsf{E}[\pi_{K,n}^N(f)]-\pi(f)\right)^2
=\left(\mathsf{E}[\pi_{n}^N(f)]-\pi(f)\right)^2
\le
\frac{\mathrm{osc}(f)^2}{n}\sum_{p=N+1}^{N+n}
\tvnorm{\delta_x \kerULA_\gamma^{p}-\pi(\cdot)}^2,
\end{equation}
Under assumptions {\bf (H1)} and {\bf (H2)}, the corresponding Markov chain has a unique stationary distribution $\pi_\gamma$, which is different from $\pi$. From \cite[Theorem~10]{durmus:moulines:2017} and Lemma~\ref{lem:v_ergodicity} it follows that
\begin{equation}
\label{eq:bias}
\tvnorm{\pi - \pi_\gamma} < C_1\sqrt{\gamma}, \quad \tvnorm{\delta_x\kerULA_{\gamma}^{n}-\pi_\gamma} \leq C_2\rho^{\gamma n}\left(V(x) + \pi_\gamma(V)\right)
\end{equation}
for  $V(x) = 1 + \|x\|^2,$ some $\rho\in (0,1)$ and constants $C_1,C_2>0$ independent of $\gamma$  and $n$. Combining \eqref{eq:bias} with the triangle inequality, we obtain \eqref{eq:sq_bias}.
\end{proof}

@@

\appendix
\section{Bounds for moments of  ULA}\label{sec:appendix}
First let us notice that conditions {\bf (H1)} and {\bf (H2)} imply (\cite[Lemma~16]{brosse2018diffusion}) that for some $K_2 > 0$ it holds
\begin{equation}
\label{eq:H3}
\ps{\nabla U(x)}{x} \geq \frac{m}{2}\|x\|^2, \quad \|x\| \geq K_2
\end{equation}

Now we shall show that conditions {\bf (H1)} and {\bf (H2)} are sufficient to show that ULA kernel $Q_\gamma$ satisfies the so-called drift condition:
\begin{definition}
\label{def:drift}
We say that the Markov kernel $Q_{\gamma}$ satisfies Foster-Lyapunov drift condition if there exist a measurable function $V: X \rightarrow [1; +\infty)$, real numbers $\lambda \in (0,1)$ and $C > 0$ such that for any $x \in X$,
\begin{equation}
\label{eq:lyapunov}
Q_{\gamma}V(x) \leq \lambda^{\gamma} V(x) + \gamma C
\end{equation}
\end{definition}

\begin{lem}
\label{lem:drift}
 Assume that the potential $U(x), x \in \mathbb{R}^d$ satisfies conditions {\bf (H1)} and {\bf (H2)} and without loss of generality consider $\nabla U(0) = 0$. Then the kernel $Q_\gamma$ from \eqref{eq:ula_kernel} satisfies drift condition \eqref{eq:lyapunov} for any $0 < \gamma < \overline{\gamma} = \frac{m}{4L}$ with drift function $V(x) = 1 + \|x\|^2$,   $\lambda = \exp{\left(-\frac{m}{2}\right)}$, $C = \frac{25K_2^2}{8} + d + m$ with $K_2$ from \eqref{eq:H3}, $m_U$ from {\bf (H2)}.
\end{lem}

\begin{proof} Let $V(x) = 1 + \|x\|^2$, then
$$
Q_{\gamma}V(x) = \int\limits_{\rset^d}V(y)Q_{\gamma}(x,dy) = \int\limits_{\rset^d}\frac{(1+\|y\|^2)}{(2\pi\gamma)^{\frac{d}{2}}}\exp{\left(-\frac{\|y-x+\gamma\nabla U(x)\|^2}{2\gamma}\right)}\,dy =
$$

$$
=1 + \frac{1}{(2\pi\gamma)^{\frac{d}{2}}}\int\limits_{\mathbb{R}^d}\|z + x - \gamma \nabla U(x)\|^2 \exp{\left(-\frac{\|z\|^2}{2\gamma}\right)}\,dz
$$
Let us first consider the case $x \notin B(0,K_2)$. Note that
$$
\|z+x-\gamma \nabla U(x)\|^2 = \|z\|^2 + 2 \langle z, x - \gamma \nabla U(x) \rangle + \|x - \gamma \nabla U(x)\|^2
$$
and the linear term vanishes after integration, moreover, for $Z \sim \mathcal{N}(0,\gamma I_d)$, it holds $\mathbb{E}\|Z\|^2 = \gamma d$. It remains to notice that due to \eqref{eq:H3},
\[
\|x - \gamma \nabla U(x) \|^2 = \|x\|^2 - 2\gamma \langle \nabla U(x), x\rangle + \gamma^2 \|\nabla U(x)\|^2 \leq \left(1 - \gamma m + 2\gamma^2 L^2\right)\|x\|^2
\]
Since $\gamma < \frac{m}{4L}$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma d + (\gamma m - 2\gamma^2 L^2) \leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma (d + m)
$$
Now let $x \in B(0,K_2)$. Then simply using $\|x-\gamma \nabla U(x)\|^2 \leq 2(1+L\gamma)^2\|x\|^2$, we obtain
$$
Q_{\gamma}V(x) \leq (1-\gamma m + 2\gamma^2 L^2)V(x) + \gamma\left((m - 2\gamma L^2)(1+\|x\|^2) + d + 2(1+L\gamma)^2\|x\|^2\right) \leq
$$

$$
\leq \exp^{-\frac{\gamma m}{2}}V(x) + \gamma(\frac{25K_2^2}{8} + d + m)
$$
\end{proof}
